---
title: 'Machine Learning 2019: Tree-Based Methods'
author: "Sonali Narang"
date: "10/28/2019"
output:
  pdf_document: default
  pdf: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Tree-Based Methods 

Decision tree is a type of supervised learning algorithm that can be used in both regression and classification problems. Tree-based methods works for both categorical and continuous input and output variables.

```{r load relevant libraries, include=FALSE}
library(tidyverse)
library(ISLR)
library(tree)
library(randomForest)
library(MASS)
library(gbm)
library(mlbench)
```
## The Carseats Dataset 

400 Observations, 11 variables
Response Variable: Sales/High 

```{r The Carseats Dataset}
data("Carseats")
carseats = Carseats
head(carseats)

#convert quantitative variable Sales into a binary response 
High = ifelse(carseats$Sales<=8, "No", "Yes")
carseats = data.frame(carseats, High)

head(carseats)
```

## Classification Tree

Input variables (X) can be continuous or categorical.
Response variable (Y) is categorical (usually binary): in this case Sales/High.

```{r Classification Tree}
#set seed to make results reproducible 
set.seed(29)

#split data into train and test subset (250 and 150 respectively)
train = sample(1:nrow(carseats), 250)

#Fit train subset of data to model 
tree.carseats = tree(High~.-Sales, carseats, subset=train)
summary(tree.carseats)

#Visualize tree
plot(tree.carseats)
text(tree.carseats, pretty=0)

#each of the terminal nodes are labeled Yes or No. The variables and the value of the splitting choice are shown at each terminal node. 

#Use model on test set, predict class labels 
tree.pred = predict(tree.carseats, carseats[-train,], type="class")

#Misclassification table to evaluate error 
with(carseats[-train,], table(tree.pred, High))

#Calculate error by summing up the diagonals and dividing by number of total predictions
mc = (71 + 42) / 150
mc
```

## Pruning using cross-validation
Pruning is a method to cut back the tree to prevent over-fitting. 

```{r Pruning}
#cross-validation to prune the tree using cv.tree
cv.carseats = cv.tree(tree.carseats, FUN = prune.misclass)

#Sizes of the trees as they were pruned back, the deviances as the pruning proceeded, and cost complexity parameter used in the process.
cv.carseats

#Visualize 
plot(cv.carseats)

#Prune tree to a size of 12 
prune.carseats = prune.misclass(tree.carseats, best = 12)

#Visualize tree 
plot(prune.carseats)
text(prune.carseats, pretty=0)

#Evaluate on test set 
tree.pred = predict(prune.carseats, carseats[-train,], type="class")

#Misclassification 
with(carseats[-train,], table(tree.pred, High))

#Error 
mc_pruning = (66 + 41) / 150
mc_pruning

##pruning did not increase misclassification error by too much and resulted in a simpler tree!!
```
Pruning did not increase misclassification error by too much and resulted in a simpler tree!!

Decision trees suffer from high variance, meaning if you split the training data into 2 parts at random, and fit a decision tree to both halves, the results that you get could be very different.

Bagging and boosting are technique used to reduce the variance of your predictions.

## The Boston Housing Dataset 

506 Observations, 14 variables
Response Variable: medv (median value of owner-occupied homes for each suburb)

```{r The Boston Housing Dataset}
data("Boston")
boston = Boston
head(Boston)
```

## Bagging: Random Forest 

Bagging involves creating multiple copies of the original training dataset using the bootstrap, fitting a separate decision tree to each copy, and then combining all of the trees in order to create a single predictive model. Each tree is built on a bootstrapped dataset, independent of the other trees.

Random Forest: Each time a split in a tree is considered, a random sample of m predictors is chosen as split candidates from the full set of p predictors. The split is allowed to use only one of those m predictors.


```{r Random Forest}
#set seed for reproducibility 
set.seed(29)

#split into train and test sets (300 and 206 respectively)
train = sample(1:nrow(boston), 300)

#fit training subset of data to model 
rf.boston = randomForest(medv~., data = boston, subset = train)
rf.boston

#summary of rf.boston gives information about the number of trees, the mean squared residuals (MSR), and the percentage of variance explained

#No. of variables tried at each split: 4 
#Each time the tree comes to split a node, 4 variables would be selected at random, then the split would be confined to 1 of those 4 variables.

##Lets try a range of mtry (number of variables selected at random at each split)
oob.err = double(13)
test.err = double(13)

#In a loop of mtry from 1 to 13, you first fit the randomForest to the train dataset
for(mtry in 1:13){
  fit = randomForest(medv~., data = boston, subset=train, mtry=mtry, ntree = 350)
  oob.err[mtry] = fit$mse[350] ##extract Mean-squared-error 
  pred = predict(fit, boston[-train,]) #predict on test dataset
  test.err[mtry] = with(boston[-train,], mean( (medv-pred)^2 )) #compute test error
}

#Visualize 
matplot(1:mtry, cbind(test.err, oob.err), pch = 23, col = c("red", "blue"), type = "b", ylab="Mean Squared Error")
legend("topright", legend = c("OOB", "Test"), pch = 23, col = c("red", "blue"))
```

## Boosting 

Boosting is another approach to improve the predictions resulting from a decision tree. Trees are grown sequentially: each tree is grown using information from previously grown trees. Each tree is fitted on a modified version of the original dataset.


```{r Boosting}
#Gradient Boosting Model
boost.boston = gbm(medv~., data = boston[train,], distribution = "gaussian", n.trees = 10000, shrinkage = 0.01, interaction.depth = 4)

#Variable Importance Plot
summary(boost.boston)

#Visualize important variables of interest
plot(boost.boston,i="lstat")
plot(boost.boston,i="rm")

#Predict on test set
n.trees = seq(from = 100, to = 10000, by = 100)
predmat = predict(boost.boston, newdata = boston[-train,], n.trees = n.trees)
dim(predmat)

#Visualize Boosting Error Plot
boost.err = with(boston[-train,], apply( (predmat - medv)^2, 2, mean) )
plot(n.trees, boost.err, pch = 23, ylab = "Mean Squared Error", xlab = "# Trees", main = "Boosting Test Error")
abline(h = min(test.err), col = "red")

```

## Homework

1. Attempt a regression tree-based method (not covered in this tutorial) on a reasonable dataset of your choice. Explain the results. 

Regression trees can be built for continuous outcomes as well, so the following is a regression tree analysis on the maximum one-hour-avg ozone value in the LA Ozone dataset.

```{r}
##define dataset and explore variables
data(Ozone)
head(Ozone)

##remove missing values
Ozone <- na.omit(Ozone)
```

```{r}
##build a regression tree to predict tax based on all other variables in the dataset
tree.model <- tree(V4 ~ ., data=Ozone)
##plot the resulting tree
plot(tree.model)
##display residuals to evaluate model quality
summary(tree.model)
```

```{r}
##build a second model by adjusting the acceptable error rate reduction from tree growth
tree.model2 <- tree(V4 ~ ., data=Ozone, mindev=0.005)
##plot the resulting tree
plot(tree.model2)
##display residuals to evaluate model quality
summary(tree.model2)
```

Decreasing the acceptable error rate reduction in tree growth predictably made for a more complex tree with more nodes. Both chose and used the same set of predictor variables. The mean squared error decreases (11.45 to 9.06) as the tree fits the data more accurately, but results in a model that may be overfit to the training data and suffer in performance on an outside dataset.

2. Attempt both a bagging and boosting method on a reasonable dataset of your choice. Explain the results.

```{r}
##define dataset and explore variables
data(Ozone)
head(Ozone)

##remove missing values
Ozone <- na.omit(Ozone)

#set seed for reproducibility 
set.seed(29)

#split into train and test sets (100 and 106 respectively)
train = sample(1:nrow(Ozone), 100)

#fit training subset of data to model 
rf.ozone = randomForest(V4~., data = Ozone, subset = train)
rf.ozone

#No. of variables tried at each split: 4 
#Each time the tree comes to split a node, 4 variables would be selected at random, then the split would be confined to 1 of those 4 variables.

##Restructure dataframe to pull outcome variable to the front
col_idx <- grep("V4", names(Ozone))
Ozone <- Ozone[, c((1:ncol(Ozone))[-col_idx], col_idx)]
names(Ozone)

##Lets try a range of mtry (number of variables selected at random at each split)
oob.err = double(12)
test.err = double(12)

#In a loop of mtry from 1 to 13, you first fit the randomForest to the train dataset
for(mtry in 1:12){
  fit = randomForest(V4~., data = Ozone, subset=train, mtry=mtry, ntree = 350)
  oob.err[mtry] = fit$mse[350] ##extract Mean-squared-error 
  pred = predict(fit, Ozone[-train,]) #predict on test dataset
  test.err[mtry] = with(Ozone[-train,], mean( (V4-pred)^2 )) #compute test error
}

#Visualize 
matplot(1:mtry, cbind(test.err, oob.err), pch = 23, col = c("red", "blue"), type = "b", ylab="Mean Squared Error")
legend("topright", legend = c("OOB", "Test"), pch = 23, col = c("red", "blue"))
```

```{r}
#Gradient Boosting Model
boost.ozone = gbm(V4~., data = Ozone[train,], distribution = "gaussian", n.trees = 10000, shrinkage = 0.01, interaction.depth = 4)

#Variable Importance Plot
summary(boost.ozone)

#Visualize important variables of interest
plot(boost.ozone,i="V1")
plot(boost.ozone,i="V2")
plot(boost.ozone,i="V3")

#Predict on test set
n.trees = seq(from = 100, to = 10000, by = 100)
predmat = predict(boost.ozone, newdata = Ozone[-train,], n.trees = n.trees)
dim(predmat)

#Visualize Boosting Error Plot
boost.err = with(Ozone[-train,], apply( (predmat - V4)^2, 2, mean) )
plot(n.trees, boost.err, pch = 23, ylab = "Mean Squared Error", xlab = "# Trees", main = "Boosting Test Error")
```

Both bagging and boosting on this dataset actually resulted in worse model performance than the baseline Random Forest that I created in the first section. The fact that both methods resulted in worse performance is probably indicative that the original model was already optimized, and that the dataset I used could not be reduced in complexity any further due to the variation of the data. The purpose of bagging is to reduce the complexity of models that overfit a training dataset, however, if the model is already optimally complex, then reducing it further results in a less accurate model. Alternatively, the purpose of boosting is to increase the complexity of a model that has high bias, but again, if the model is already optimized, then increasing the complexity will result in worse performance.